---
title: "STAT184 Final Project"
author: "Abhiraj Srivastava"
format: pdf
editor: visual
toc: true
number-sections: true
---

# Introduction

This project explores trends in public confidence toward medicine using data from the General Social Survey (GSS). The analysis focuses on how trust in medical institutions changed between 2018 and 2024, particularly before and after the COVID-19 pandemic. Public trust in medicine is an important social indicator because it affects healthcare behavior, vaccine uptake, and confidence in scientific institutions.

# Data Provenance

The dataset used in this project comes from the General Social Survey (GSS), a nationally representative survey conducted in the United States. The GSS collects demographic, behavioral, and attitudinal data from respondents across the country.

The specific variable used in this analysis is:

- `CONMEDIC` — confidence in medicine

Additional variables from the GSS may also be used for exploratory analysis and comparison.

# FAIR and CARE Principles

## FAIR Principles

The dataset partially satisfies the FAIR principles:

- **Findable:** The GSS dataset is publicly available and well documented.
- **Accessible:** Researchers and students can access the data online.
- **Interoperable:** The dataset works with statistical software such as R.
- **Reusable:** Documentation and variable descriptions allow future reuse.

## CARE Principles

The CARE principles are only partially satisfied:

- The GSS provides broad demographic representation.
- However, the data are not specifically designed around Indigenous data governance or community ownership principles.

# Data Cleaning and Wrangling

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyr)
library(kableExtra)
library(dplyr)
library(ggradar)
library(fmsb)
library(scales)

GSS_clean <- readRDS("GSS_clean.rds")
```

The data required cleaning because missing values in the GSS are often represented with negative numbers such as `-100`. These values were converted into `NA` values before analysis.

```{r}
medical_data <- GSS_clean |>
  filter(YEAR %in% c(2018, 2021, 2022, 2024)) |>
  mutate(
    CONMEDIC = if_else(CONMEDIC < 0, NA_integer_, CONMEDIC)
  )
```

# Exploratory Data Analysis

The exploratory analysis focuses on trends in medical confidence across multiple years.

## Fig 1.1, Medical Confidence Table

```{r}
medical_table <- medical_data |>
  mutate(
    Medical_Confidence = case_when(
      CONMEDIC == 1 ~ "High Trust / Full Confidence",
      CONMEDIC == 2 ~ "Moderate / Skeptical",
      CONMEDIC == 3 ~ "Low / No Confidence",
      TRUE ~ NA_character_
    )
  ) |>
  drop_na(Medical_Confidence) |>
  count(YEAR, Medical_Confidence) |>
  group_by(YEAR) |>
  mutate(
    Percent = round(n / sum(n) * 100, 1)
  ) |>
  ungroup()

knitr::kable(
  medical_table,
  caption = "Medical Confidence by Year",
  align = "c"
)
```


## Figure 1.2, Stacked Bar Chart of Medical Confidence

```{r}
medical_graph_data <- medical_data |>
  mutate(
    Sentiment = case_when(
      CONMEDIC == 1 ~ "High Trust / Full Confidence",
      CONMEDIC == 2 ~ "Moderate / Skeptical",
      CONMEDIC == 3 ~ "Low / No Confidence",
      TRUE ~ NA_character_
    )
  ) |>
  drop_na(Sentiment) |>
  count(YEAR, Sentiment) |>
  group_by(YEAR) |>
  mutate(percent = n / sum(n)) |>
  ungroup() |>
  mutate(
    Sentiment = factor(
      Sentiment,
      levels = c(
        "High Trust / Full Confidence",
        "Moderate / Skeptical",
        "Low / No Confidence"
      )
    )
  )

ggplot(
  medical_graph_data,
  aes(
    x = factor(YEAR),
    y = percent,
    fill = Sentiment
  )
) +
  geom_col(
    width = 0.9,
    color = "white"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%"),
    limits = c(0, 1)
  ) +
  scale_fill_manual(
    values = c(
      "Low / No Confidence" = "#e45756",
      "Moderate / Skeptical" = "#f2f2f2",
      "High Trust / Full Confidence" = "#1f77b4"
    )
  ) +
  labs(
    title = "The Shift in Public Sentiment Toward Medicine",
    subtitle = "Percentage of population by trust level (2018–2024)",
    x = NULL,
    y = "Percentage of Respondents",
    fill = "Sentiment"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 20
    ),
    plot.subtitle = element_text(size = 14),
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

\`\`\`{r}\
\# R code goes here\
summary(cars)\
\`\`\`

## Interpretation of Results

The visualization suggests that public confidence in medicine shifted noticeably after 2020. The percentage of respondents expressing low confidence increased over time, while high trust gradually declined. This trend may reflect the political and social polarization surrounding public health during and after the COVID-19 pandemic.

## Figure 2, Gradient Table of Confidence, Happiness, and Prayer

```{r}
# make a vector with excluded values (defined in the social survey codebook)
exclude_vals <- c(-100, -99, -98, -97, -96, -95, -94, -93, -90, -80, -70, -60, -40)

#get averages
summary_table <- GSS_clean |>
  
#filter for specified years
  filter(YEAR %in% c(2018, 2021, 2022, 2024)) |>
  mutate(across(c(PRAY, HAPPY, CONCLERG, CONFED, CONMEDIC),
                ~ ifelse(. %in% exclude_vals, NA, .))) |>
#calculate means (prayer divided by two to match the 1-3 scores)
   group_by(YEAR) |>
  summarise(
    Prayer = mean(PRAY, na.rm = TRUE) / 2,
    Happiness = mean(HAPPY, na.rm = TRUE),
    `Clergy Confidence` = mean(CONCLERG, na.rm = TRUE),
    `Govt Confidence` = mean(CONFED, na.rm = TRUE),
    `Health Confidence` = mean(CONMEDIC, na.rm = TRUE)
  ) |>
  pivot_longer(cols = -YEAR, names_to = "variable", values_to = "mean") |>
  pivot_wider(names_from = YEAR, values_from = mean)

#Create a separate dataframe with the formatted cells
formatted_table <- summary_table

#make objects with gradient values
green_anchor <- 1.43
white_anchor <- 2.38

#loop through each cell in the table and set the color value relative to the gradient values 
for(i in 1:nrow(summary_table)) {
  for(col in c("2018", "2021", "2022", "2024")) {
    
    val <- summary_table[[col]][i]
    
    if(!is.na(val) && is.numeric(val)) {
      if(val <= green_anchor) {
        intensity <- 1 #white
      } else if(val >= white_anchor) {
        intensity <- 0 #green
      } else {
        intensity <- 1 - ((val - green_anchor) / (white_anchor - green_anchor))
      }
      intensity <- max(0, min(1, intensity))
#set intensity values to actual RGB codes
      red_blue_value <- round(255 - (intensity * 255))
      green_value <- 255
      bg_color <- rgb(red_blue_value, green_value, red_blue_value, maxColorValue = 255)

#text readability
      formatted_table[[col]][i] <- cell_spec(round(val, 2), 
                                             background = bg_color,
                                             color = ifelse(intensity > 0.6, "white", "black"))
    } else {
      formatted_table[[col]][i] <- cell_spec("NA", background = "#f0f0f0", color = "gray")
    }
  }
}

#final kable table
styled_table <- formatted_table |>
  kable(format = "html", 
        caption = "<b style='color:black;'>Table 1: Mean Responses by Variable and Year</b>", 
        
        escape = FALSE, 
        align = "c") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                font_size = 14) |>
  row_spec(0, bold = TRUE, background = "#f0f0f0") |>
  footnote(general = paste(
    "Note: Variable scales - PRAY: 1 (Several times/day) to 6 (Never), divided by 2 for 0.5-3 scale;",
    "HAPPY: 1 (Very happy) to 3 (Not too happy);",
    "CONCLERG: 1 (A great deal of confidence in clergy) to 3 (Hardly any);",
    "CONFED: 1 (A great deal of confidence in federal government) to 3 (Hardly any);",
    "CONMEDIC: 1 (A great deal of confidence in medicine) to 3 (Hardly any).",
    "Green gradient: Darker green = closer to 1 (more positive response), White = closer to 3 (more negative response)."
  ))
```
## Interpretation of Results

This table shows us a couple interesting things. Number one, the gradient scale needed to be adjusted to a smaller scale than 1-3 in order to show contrast, meaning that the data were not very diffuse. Additionally, trust in government was particularly low compared to other metrics.

##Figure 3.1 Radar Chart of Confidence, Happiness, and Prayer

```{r}
#Wrangling, this section is adapted from Gradient-table-code.R (variable names are shortened)
exclude_vals <- c(-100, -99, -98, -97, -96, -95, -94, -93, -90, -80, -70, -60, -40)

summary_table <- GSS_clean |>
  filter(YEAR %in% c(2018, 2021, 2022, 2024)) |>
  mutate(across(c(PRAY, HAPPY, CONCLERG, CONFED, CONMEDIC),
                ~ ifelse(. %in% exclude_vals, NA, .))) |>
  group_by(YEAR) |>
  summarise(
    Prayer = mean(PRAY, na.rm = TRUE) / 2,
    Happiness = mean(HAPPY, na.rm = TRUE),
    `Clergy Conf.` = mean(CONCLERG, na.rm = TRUE),    
    `Govt. Conf.` = mean(CONFED, na.rm = TRUE),     
    `Health Conf.` = mean(CONMEDIC, na.rm = TRUE) 
  ) |>
  pivot_longer(cols = -YEAR, names_to = "variable", values_to = "mean") |>
  pivot_wider(names_from = YEAR, values_from = mean)

#Transform dataframe so years are the rows
transformed_summary_table <- summary_table |>
  pivot_longer(cols = -variable, names_to = "year", values_to = "value") |>
  pivot_wider(id_cols = year, names_from = variable, values_from = value)

#Single Comparison Chart
radar_data <- transformed_summary_table
rownames(radar_data) <- radar_data$year
radar_data <- radar_data[, -1] #take year column out

#set scale and margins
radar_data <- rbind(
  rep(2.40, ncol(radar_data)),
  rep(1.40, ncol(radar_data)),
  radar_data
)
par(mar = c(1, 1, 2, 1))

#comparison chart
radarchart(radar_data,
           axistype = 1,
           pcol = c("#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4"),
           pfcol = c(rgb(255,107,107, alpha=100, maxColorValue=255),
                     rgb(78,205,196, alpha=100, maxColorValue=255),
                     rgb(69,183,209, alpha=100, maxColorValue=255),
                     rgb(150,206,180, alpha=100, maxColorValue=255)),
           plwd = 3,
           cglcol = "grey",
           cglty = 1,
           cglwd = 0.8,
           axislabcol = "black",
           title = "Survey Responses by Year",
           caxislabels = c("1.40", "1.65", "1.90", "2.15", "2.40"),
           vlcex = .6,        # Variable labels (Clergy Conf., etc.) - 1.2x larger
           cex.main = 1,     # Title size - 1.8x larger
           calcex = .4,       # Axis numbers (1.40, 1.65, etc.) - 0.9x
           
        )

# add legend (made smaller to not conflict with labels)
legend(x = 1.7, y = 1.15,
       legend = transformed_summary_table$year,
       fill = c("#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4"),
       title = "Year",
       cex = 0.5,
       pt.cex = 0.5,
       title.cex = 0.5,
       bty = "n",
       xpd = TRUE)
par(mar = c(.5, .4, .4, .2) + 0.1)
```

##Figure 3.2 Radar Chart of Confidence, Happiness, and Prayer By Year

```{r}

#four seperate charts

#set a grid
par(mfrow = c(2, 2))
par(mar = c(1, 1, 3, 1))

#set colors to years
year_colors <- list(
  "2018" = "#FF6B6B",
  "2021" = "#4ECDC4",
  "2022" = "#45B7D1",
  "2024" = "#96CEB4"
)

#Create seperate radar charts with loops
for(year_name in transformed_summary_table$year) {
  year_data <- transformed_summary_table |>
    filter(year == year_name) |>
    select(-year)
  
  year_matrix <- rbind(
    rep(2.40, ncol(year_data)),
    rep(1.40, ncol(year_data)),
    year_data
  )

  #radar charts
  radarchart(year_matrix,
             axistype = 1,
             pcol = year_colors[[year_name]],
             pfcol = adjustcolor(year_colors[[year_name]], alpha.f = 0.4),
             plwd = 3,
             cglcol = "grey",
             cglty = 1,
             cglwd = 0.8,
             axislabcol = "black",
             vlcex = 1.1,
             title = paste("Year:", year_name),
             caxislabels = c("1.40", "1.65", "1.90", "2.15", "2.40"))
}

#fix layout
par(mfrow = c(1, 1))
par(mar = c(.5, .4, .4, .2) + 0.1)
```

##Interpretation

This particular visualization method shows a different story from the gradient table, using this method, we see that each radar chart shows a similar shape throughout each year, this is evident when all four radar charts are overlayed.

# Conclusion

This analysis demonstrated how confidence in medicine changed between 2018 and 2024 using data from the General Social Survey. Through data cleaning, exploratory analysis, tables, and visualizations, the report highlighted changing public sentiment toward medical institutions in the United States.

# Code Appendix

```{r}
sessionInfo()
```

{r}
